Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-01-19 paper/paper.tex
second results subsection

The expected counts (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figs/corr.png}
\caption{
Correlations between amino-acid mutation fitness effect estimates made using {\bf (A)} different viral clades or {\bf (B)} sequences from only the USA or only England.
Each point is an amino-acid mutation, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html} for interactive plots that include additional viral clades, allow adjustment of the expected-count cutoff, and allow examination of just specific genes.
\label{fig:corr}
}
\end{figure*}

The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
The intuition is straightforward: mutations arise at all sites during viral replication, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations scales roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends providing links to interactive plots that enable adjustment of this threshold.

The estimated effects of mutations are robust to partitioning

\subsection*{new subsection}

@@

>>>>

The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.

@@

The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
The intuition is straightforward: mutations arise at all sites during viral replication, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations scales roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends linking to interactive plots that enable adjustment of this threshold.

\subsection*{Mutation-effect estimates are robust to subsampling natural sequences, with some evidence of epistasis in spike}

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figs/corr.png}
\caption{
Correlations of mutation fitness effect estimates made using subsets of natural sequences.
Correlations between estimates made {\bf (A)} just using sequences from the Delta or Omicron BA.5 clades or {\bf (B)} just from the USA or England.
Each point is an amino-acid mutation, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
Only mutations with at least 10 expected counts are shown, which is why panels have different numbers of mutations shown (sequence subsets vary in size).
{\bf (C)} Correlations between clade or geography subsets become higher with an increasingly large threshold for minimum expected counts.
Spike mutations have a worse correlation when subsetting by viral clade (plot shows average correlation over all pairwise combinations of Delta, BA.1, BA.2, and BA.5), but not when subsetting by geography (USA or England).
{\bf (D)} Correlations in estimated mutation effects decline for clades with higher protein divergence, with the effect most noticeable for spike since spike is more diverged among SARS-CoV-2 clades than other viral proteins.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html} for versions of A and B that include all viral clades with at least 500,000 total expected counts (summed across all mutations) and have other interactive options.
\label{fig:corr}
}
\end{figure*}

To assess the robustness of the estimated effects of individual amino-acid mutations, we computed the correlations among estimates made using subsets of SARS-CoV-2 sequences from different viral clades or geographic locations.
These estimates were highly correlated, although there is modest variation in estimates across sequence subsets (Figure~\ref{fig:corr}A,B).

The modest variation in mutation effects estimated from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real differences in effects of mutations among subsets of SARS-CoV-2.
It is not obvious why there should be real differences in mutation effects in geographic locations, but there is experimental evidence that epistasis has shifted effects of some spike mutations during viral evolution~\citep{starr2022shifting,moulana2022compensatory,mccallum2021n}.
To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
Correlations increased with this count threshold, consistent with a reduction in statistical noise when using larger mutation counts.
But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
The lower cross-clade correlation for spike appears to be due to epistatic shifts in mutation effects that arise as sequences diverge~\citep{starr2022shifting, pollock2012amino, shah2015contingency}, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
However, for the rest of this paper we aggregate counts across clades, since for most purposes the benefits aggregating counts (which enables lower-noise estimates more mutations) outweighs the effect of modest epistatic shifts.

\subsection*{Structural and non-structural proteins are under strong purifying selection, but many accessory genes are not}

@@

Caveats include synonymous selection.
